Search for WIMP Inelastic Scattering Off Xenon Nuclei With XENON100 Data

### 2017-01-18 Inelastic/analysis.tex
updated_intro_and_ref
updated_xenon100_started_analysis_section

\section{Data Analysis}

This analysis is performed using XENON100 Run-II science data, which corresponds to an exposure of 224.6~live days. The detector response to
electronic recoil (ER) has been characterized using $^{60}$Co and $^{232}$Th radioactive sources, while response to inelastic nuclear recoil (NR)
scattering was calibrated using an $^{241}$AmBe source.

The inelastic scattering of a WIMP with the nucleus of $^{129}$Xe produces an energy deposit via nuclear recoil with subsequent emission of  
a 39.6~keV de-excitation photon. 
The largest fraction of the energy released in the event is via electronic recoil due to the emitted photon, this represents an
unusual signature for this kind of detector and brings the possible signal to overlay a phase space region with large backgrounds.
The chosen region of interest (ROI) for this analysis surrounds the 39.6~keV xenon line in the cS1-cS2 plane and is further divided into
sub-regions, as shown in Figure~\ref{fig:SR} and~\ref{fig:SR2}.

Events are asked, other than falling in the defined region of interest, to fulfill several selection criteria which can be summarized as:
selection aimed to reduce noise impact and including energy (S1) and S2 thresholds, veto of events with energy release in the detector's outer volume, 
and additionally, events must be of single scatter nature and fall into a predefined fiducial volume. This analysis follows the 
selection criteria described in detail in \cite{dataAnalysis} for Run-II, with only few exceptions reported in what follows. 
%only few modification have been designed specifically for this analysis and discussed below. 
The selection on S2 width as a function of drift time has been optimized on a sample of events selected from the 39.6~keV line
and set to a 95\% acceptance on these. Events are required to be single scatter by applying a threshold on the 
second largest S2 peak size,  for this analysis this threshold has been optimized to 160~PE and set constant as function of S2 signal size. 
Finally the chosen fiducial volume corresponds to 34~Kg of liquid xenon.

@@

\subsection{Signal Simulation} 
%description of the simulated signal, few words about cross checks MC matching.
The detector response to inelastic scattering of WIMPs  off $^{129}$Xe nucleus was simulated using an empirical model.
The total deposited energy is divided into two independent contributions: one related to the 39.6~keV de-excitation photon and the other  relative to
the simultaneous nuclear recoil of the xenon atom. The number of photons and charge yield detected is simulated separately for each contribution
and then added together.

>>>>

\section{Data Analysis}
\label{sec:analysis}

This analysis is performed using XENON100 run\,II science data, which corresponds to a data set with an exposure of 224.6~live days. The detector response to
ERs has been characterised with $^{60}$Co and $^{232}$Th calibration sources, while the response to NRs was calibrated with an $^{241}$AmBe $(\alpha,n)$-source. This fast neutron source gives rise to elastic and inelastic neutron-nucleus scatters, and can thus be employed to define the expected signal region.

\subsection{Signal Correction}

A particle interaction in the liquid xenon produces an S1 and a correlated S2 signal with a certain number of photoelectrons (PE) observed by the PMTs. The non-uniform scintillation light collection by the PMT arrays, due to solid angle effects, Rayleigh scattering length, reflectivity, transmission of the electrodes, etc, lead to a position-dependent S1 signal. The warping of the top meshes (inducing a variation in the width of the gas gap between the anode and the liquid-gas interface), the absorption of electrons by residual impurities, as they drift towards the gas region, as well as solid angle effects lead to a position-dependent S2 signal. These signals are thus corrected in 3 dimensions, using various calibration data~\cite{Aprile:2011dd}, with the corrected quantities denoted as cS1 and cS2.

\subsection{Signal Region}

As explained in Section~\ref{sec:intro}, the inelastic scattering of a WIMP with a $^{129}$Xe nucleus produces an energy deposit via a NR with subsequent emission of  
a 39.6\,keV de-excitation photon. The largest fraction of the energy released in the event is via an ER, due to the emitted photon which loses its energy in the LXe.
This represents an unusual signature compared to the one expected from an elastic scatter, and brings the signal region to overlap with the ER background region.
The selected region of interest (ROI) for this analysis surrounds the 39.6\,keV xenon line in the cS1-cS2 plane and is further divided into
sub-regions, as shown in Figures~\ref{fig:SR} and~\ref{fig:SR2}.

Other than falling in the defined region of interest, events are required to fulfil several selection criteria, which can be summarised as follows:
selection aimed to reduce noise impact and including energy (S1) and S2 thresholds, veto of events with energy release in the detector's outer volume, 
and additionally, events must be of single scatter nature and fall into a predefined fiducial volume. This analysis follows the 
selection criteria described in detail in \cite{dataAnalysis} for Run-II, with only few exceptions reported in what follows. 
%only few modification have been designed specifically for this analysis and discussed below. 
The selection on S2 width as a function of drift time has been optimized on a sample of events selected from the 39.6~keV line
and set to a 95\% acceptance on these. Events are required to be single scatter by applying a threshold on the 
second largest S2 peak size,  for this analysis this threshold has been optimized to 160~PE and set constant as function of S2 signal size. 
Finally the chosen fiducial volume corresponds to 34~Kg of liquid xenon.

@@

\subsection{Signal Simulation} 
%description of the simulated signal, few words about cross checks MC matching.

The detector response to inelastic scattering of WIMPs  off $^{129}$Xe nucleus was simulated using an empirical model.
The total deposited energy is divided into two independent contributions: one related to the 39.6~keV de-excitation photon and the other  relative to
the simultaneous nuclear recoil of the xenon atom. The number of photons and charge yield detected is simulated separately for each contribution
and then added together.

### 2017-01-18 Inelastic/intro.tex
updated_xenon100_started_analysis_section

\section{\label{sec:level1} Introduction}
Astrophysical and cosmological evidence indicates that the dominant mass fraction of our Universe consists of some yet unknown
form of dark, or invisible matter. The dark matter could be made of new, yet undiscovered particles, and well-motivated
theoretical models going beyond the Standard Model of particle physics predict the existence of Weakly Interacting Massive
Particles (WIMPs), which are natural candidates for dark matter. This hypothesis is currently being tested by several direct
and indirect detection experiments, as well as at the LHC~\cite{Bertone:2010zza,Baudis:2016qwx}.

@@

The XENON100 experiment operated a dual-phase (liquid and gas) xenon time projection chamber (TPC) at the Laboratori Nazionali 
del Gran Sasso (LNGS) in Italy. It contained 161\,kg of xenon in total,  with  62\,kg in the active region of the TPC. These 
were monitored by 242 1-inch square, low-radioactivity, UV-sensitive photomultiplier tubes (PMTs) arranged in two arrays, one in the liquid 
and one in the gas. The PMTs detected the prompt scintillation (S1) and the delayed, proportional scintillation signal (S2) 
created by a particle interacting in the active TPC region. The S2-signal was generated due to ionisation electrons, drifted 
in a field of 530\,V/cm and extracted into the gas phase by a field of 12\,kV/cm, where the proportional scintillation, or electroluminiscence, 
was produced. These photons carried the $x-y$ information of the interaction site, while the $z-$information came from the drift time measurement. The TPC 
thus yielded a three-dimensional event localisation, enabling to reject the majority of background events via fiducial volume 
cuts, as well a multiple interactions, which are unlikely due to dark matter particles~\cite{Aprile:2011dd}. The ratio of 
S2/S1 provided the basis for distinguishing between nuclear recoils (NRs), as induced by fast neutrons and expected from 
elastic WIMP-nucleus scatters, and electronic recoils (ERs) produced by $\beta$ and $\gamma$-rays. As we shall see, the 
inelastic scattering signature will be a superposition of an ER and a NR, due to the low lifetime of the excited state and 
the short mean free path ($\sim$0.15\,mm) of the 39.6\,keV photon.

XENON100 has acquired science data between 2008-2015, and has set competitive constraints on spin-independent~\cite{Aprile:2012nq,Aprile:2016swn} 
and spin-dependent~\cite{Aprile:2013doa,Aprile:2016swn} elastic WIMP-nucleus 
scatters, as well as on axions/ALPs~\cite{Aprile:2014eoa} and leptophilic dark matter models~\cite{Aprile:2015ade,Aprile:2015ibr,Aprile:2017yea}.

Here

>>>>

\section{\label{sec:intro} Introduction}

Astrophysical and cosmological evidence indicates that the dominant mass fraction of our Universe consists of some yet unknown
form of dark, or invisible matter. The dark matter could be made of new, stable or long-lived and yet undiscovered particles. Well-motivated
theoretical models going beyond the Standard Model of particle physics predict the existence of Weakly Interacting Massive
Particles (WIMPs), which are natural candidates for dark matter. This hypothesis is currently being tested by several direct
and indirect detection experiments, as well as at the LHC~\cite{Bertone:2010zza,Baudis:2016qwx}.

@@

The XENON100 experiment operates a dual-phase (liquid and gas) xenon time projection chamber (TPC) at the Laboratori Nazionali 
del Gran Sasso (LNGS) in Italy. It contains 161\,kg of xenon in total,  with  62\,kg in the active region of the TPC. These 
are monitored by 242 1-inch square, low-radioactivity, UV-sensitive photomultiplier tubes (PMTs) arranged in two arrays, one in the liquid 
and one in the gas. The PMTs detect the prompt scintillation (S1) and the delayed, proportional scintillation signal (S2) 
created by a particle interacting in the active TPC region. The S2-signal is generated due to ionisation electrons, drifted 
in an electric field of 530\,V/cm and extracted into the gas phase by a stronger field of $\sim$ 12\,kV/cm, where the proportional scintillation, or electroluminiscence, 
is produced. These photons carry the $(x,y)$ information of the interaction site, while the $z-$information comes from the drift time measurement. The TPC 
thus yields a three-dimensional event localisation, with an $(x,y)$ resolution of $<$3\,mm (1\,$\sigma$), and a $z$ resolution of  $<$0.3\,mm (1\,$\sigma$), enabling to reject the majority of background events via fiducial volume cuts~\cite{Aprile:2011dd}. The ratio S2/S1 provides the basis for distinguishing between nuclear recoils (NRs), as induced by fast neutrons and expected from elastic WIMP-nucleus scatters, and electronic recoils (ERs) produced by $\beta$ and $\gamma$-rays.

XENON100 has acquired science data between 2008-2015, and has set competitive constraints on spin-independent~\cite{Aprile:2012nq,Aprile:2016swn} 
and spin-dependent~\cite{Aprile:2013doa,Aprile:2016swn} elastic WIMP-nucleus 
scatters, on solar axions and galactic ALPs~\cite{Aprile:2014eoa}, as well as on leptophilic dark matter models~\cite{Aprile:2015ade,Aprile:2015ibr,Aprile:2017yea}.

Here we explore a potential new signature in the XENON100 detector, caused by spin-dependent, inelastic WIMP-$^{129}$Xe scatters. The expected  
inelastic scattering signature is a combination of an ER and a NR, due to the short lifetime of the excited nuclear state and  the short mean free path of $\sim$0.15\,mm of the 39.6\,keV de-excitation photon.
